---
title: "class18"
author: "Dalena (A17327787)"
format: pdf
toc: TRUE
---

## Background

Pertussis (a.ka. whopping cough) is a common lung infection caused by the bacteria *B. Pertussis*.

The CDC tracks cases of Pertussis in the US: https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html

## Investing Pertussis Cases By Year

```{r}
cdc <- data.frame( year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L,2022L,2023L,2024L),

cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116,3044,7063,35493)
       )
```



> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called `cdc` and use ggplot to make a plot of cases numbers over time.

```{r}
library(ggplot2)
```

```{r}
ggplot(cdc) +
  aes(year, cases) +
  geom_point() + 
  geom_line()
```

## A tale of two vaccines (wP & aP)

> Q2. Using the ggplot `geom_vline()` function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). 2020 is the CMI-PB Project. WHat do you notice?

```{r}
library(ggplot2)
ggplot(cdc) +
  aes(year, cases) +
  geom_point() + 
  geom_line() +
geom_vline(xintercept = 1946, col="dodgerblue3") +
  geom_vline(xintercept = 1996, col="red4") +
  geom_vline(xintercept = 2020, col="seagreen3")
```

>> The newer **aP** is less effective in long term than the older **wP** vaccine.

## Exploring CMI-PB data

CMI-PB (Computational Models of Immunity - Pertussis boost). The new and ongoing CMI-PB project aims to provide the scientific community with this very information. In particular, CMI-PB tracks and makes freely available long-term humoral and cellular immune response data for a large number of individuals who received either DTwP or DTaP combination vaccines in infancy followed by Tdap booster vaccinations. This includes complete API access to longitudinal RNA-Seq, AB Titer, Olink, and live cell assay results directly from their website: https://www.cmi-pb.org/



Use **jsonlitee** package

```{r}

##Read, write and process JSON data

library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject", 
                     simplifyVector = TRUE)

head(subject)
```

> Q3. How many subjects (i.e. enrolled people) are there in this dataset?

```{r}
nrow(subject)

```

> Q4. How many "aP" and "wP" infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

> Q5. How many male/female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex?

```{r}
table(subject$race, subject$biological_sex)
```

> Q7. Is this representative of the US Population?

>> No.

Let's read another database table from the CMI-PB

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen",
                      simplifyVector = TRUE)

ab_data <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer",
                     simplifyVector = TRUE)
```


"Join" these tables to get our information together. Use **dplyr** package and the `inner_join()` function.

```{r}
library(dplyr)
meta <- inner_join(subject, specimen)

head(meta)
```

```{r}
head(ab_data)
```

"join" again to get ab_data and meta all together
```{r}
abdata <- inner_join(ab_data, meta)
head(abdata)
```
```{r}
dim(abdata)
```

> Q8. How many Ab isotypes are there in the dataset?

```{r}
table(abdata$isotype)
```

> Q9. How many different antigens are measured in the dataset?

```{r}
table(abdata$antigen)
```

> Q10. Make a boxplot of antigen levels across the whole dataset (MFI vs antigen)

```{r}
ggplot(abdata) +
  aes(MFI, antigen) +
  geom_boxplot()
```

 > Q11. Are there obvious differences between aP and wP values?
 
```{r}
ggplot(abdata) +
  aes(MFI, antigen, col=infancy_vac) +
  geom_boxplot()
```
 
## Focus on IgG levels
 
IgG is the most abundant antibody in blood with four subclasses
 
```{r}
igg <- abdata |> 
  filter(isotype == "IgG")

head(igg)
```
 
Same boxplot of antigen as before
 
```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(~visit)
```

Focus in further in just one of these antigens - let's pick PT (Pertussis Toxin, one of the main toxins of the bacteria)

**2021_dataset** again for **IgG** antibody isotopes

```{r}
table(igg$dataset)
```
 
```{r}
pt_igg <- abdata |>
  filter(isotype=="IgG", 
         antigen=="PT", 
         dataset=="2021_dataset")
```
 
```{r}
dim(pt_igg)
```
 
```{r}
ggplot(pt_igg) + 
  aes(actual_day_relative_to_boost,
      MFI_normalised, 
      col=infancy_vac,
      group=subject_id) +
  geom_point() +
  geom_line() +
  theme_bw() +
  geom_vline(xintercept = 0) +
  geom_vline(xintercept = 14)

```
 

