---
title: "class17"
author: "Dalena (A17327787)"
format: pdf
---
## Downstream Analysis

Use `tximport()` by installing the **tximport** package enabling straightforward import of Kallisto results

```{r}
library(tximport)

folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path( folders, "abundance.h5" )
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```

```{r}
head(txi.kallisto$counts)
```

> Q. How many transcripts for each sample?

```{r}
colSums(txi.kallisto$counts)
```

>Q. How many transcripts are detected in at least one sample?

```{r}
sum(rowSums(txi.kallisto$counts)>0)
```

Filter annotated transcripts with no reads and those with no change over the samples

```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]

keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```

## Principal Component Analysis

```{r}
pca <- prcomp(t(x), scale=TRUE)

summary(pca)
```

### Plots
```{r}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
```

Using ggplot

```{r}
library(ggplot2)
library(ggrepel)
```

```{r}
## Make metadata
colData <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(colData) <- colnames(txi.kallisto$counts)
```

```{r}
## Make data.frame & plot

y <- as.data.frame(pca$x)
y$Condition <- as.factor(colData$condition)

ggplot(y) +
  aes(PC1, PC2, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```

## Differential-expresion analysis (DESeq2)

```{r}
library(DESeq2)
```

```{r}
sampleTable <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(sampleTable) <- colnames(txi.kallisto$counts)
```

```{r}
dds <- DESeqDataSetFromTximport(txi.kallisto,
                                sampleTable, 
                                ~condition)
```

```{r}
dds <- DESeq(dds)
```

```{r}
res <- results(dds)
head(res)
```

